\documentname: Is machine learning good or bad for the natural sciences?

### 2024-05-23 paper/main.tex
nearly camera ready

% To do:
% ------
% - consider taking one non-trivial but simple-to-understand example and thread it through
% - Take feedback from Margossian (email 2024-01-03).
% - Take feedback from Rix comments (email 2024-01-04-ish).
% - Take feedback from KSF comments (email 2024-01-19-ish).
% - Take feedback from excellent Contardo comments (email 2024-01-28).
% - Take feedback from Foit email (2024-01-29)
% - submit to ICML.

% To do after ICML submission:
% ----------------------------
% - audit bibliography; make sure arXiv references in the reference list are consistent? Also capitalization!
% - Update appendix with feedback from Vedant Chandra (email). In particular, modify model complexity and information content.
% - Should there be another BAD thing, which is about statistics for comparison of simulations and data? This is a great one; consider after ICML submission.
% - Publish on the arXiv.

% Style notes:
% ------------
% - Be very careful about the SUBJECT of sentences. Is it ML that is bad? Is it the practice that is bad? Is it the practitioner that is bad?
% - Should we use NS for natural science?
% - How many authors are there? Should we switch to first-person singular?
% - Audit for flexible, expressive, high in capacity, and so on.
% - [more here]

% HOGG CONDITIONAL
\newif\ifarxiv
\arxivfalse % set to true for arXiv submission and do something way more complicated for final ICML camera ready.

% NOW it's GO TIME
\ifarxiv
\documentclass[10pt]{article}
\else
\documentclass{article}
\fi

@@

@@

% Hogg typesetting issues that won't conflict with ICML
\frenchspacing
\newcommand{\documentname}{\textsl{Position Paper}}
\newcommand{\draftstatus}{\textcolor{red}{\sffamily\slshape Draft version 2024-01-11}}
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\newcommand{\figref}[1]{\figurename~\ref{#1}}
\newenvironment{hoggnumerate}
  {\begin{list}
    {\arabic{enumii}.}
    {\usecounter{enumii}
     %\setlength{\labelwidth}{3em}
     %\setlength{\labelsep}{0em}
     \setlength{\itemsep}{0in}
     \setlength{\parsep}{0in}
     \setlength{\parskip}{0in}
     %\setlength{\leftmargin}{1.5cm}
     %\setlength{\rightmargin}{2cm}
     %\setlength{\itemindent}{0em} 
     %\let\makelabel=\makeboxlabel
    }
  }
{\end{list}}

% Hogg typesetting issues that will conflict with ICML
\ifarxiv
\setlength{\textwidth}{5.50in}
\setlength{\oddsidemargin}{-0.25in}
\setlength{\textheight}{9.25in}
\setlength{\topmargin}{-0.50in}
\setlength{\parindent}{1.2\baselineskip} % seriously
\renewcommand{\title}[1]{\section*{\raggedright #1}}
\renewcommand{\author}[1]{\medskip\par\noindent\textbf{#1}}
\newcommand{\affil}[1]{{\footnotesize\par\noindent #1 \par}}
\renewenvironment{abstract}{\paragraph{Abstract:}}{}
\pagestyle{myheadings}
\markboth{foo}{\textcolor{gray}{\textsf{Hogg / Is machine learning good or bad for the natural sciences? \draftstatus}}}
\linespread{1.08}
\raggedbottom\sloppy\sloppypar
\else
\fi

@@

\icmltitlerunning{Is machine learning good or bad for the natural sciences?}
\begin{document}
\twocolumn[
\icmltitle{Is machine learning good or bad for the natural sciences?}

\begin{icmlauthorlist}
\icmlauthor{David W. Hogg}{ccpp,cds,mpia,flatiron}
\end{icmlauthorlist}
\icmlaffiliation{ccpp}{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726~Broadway, New~York,~NY 10003, USA}
\icmlaffiliation{cds}{Center for Data Science, New York University, 60 Fifth Avenue, New York, NY 10011, USA}
\icmlaffiliation{mpia}{Max-Planck-Institut f{\"u}r Astronomie, K{\"o}nigstuhl 17, D-69117 Heidelberg, Germany}
\icmlaffiliation{flatiron}{Flatiron Institute, a division of the Simons Foundation, 162 Fifth Avenue, New~York,~NY 10010, USA}
\icmlcorrespondingauthor{David W. Hogg}{david.hogg@nyu.edu}

@@

\vskip 0.3in
]

\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here we identify locations for ML in the natural sciences where the ontology and epistemology are valuable.
  For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can sometimes make the results more trustworthy.
  We also show that there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
  For one, when ML models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint analyses without taking on uncontrolled biases.
  The question of the title is being asked of all of the natural sciences, but the answers here are from the perspective of physics.
\end{abstract}

\section{Introduction}\label{sec:intro}
It is an understatement to say that machine learning (ML) is having a big impact across the sciences.
A significant fraction of all scientific papers in the natural sciences now employ ML in part (or all) of their analyses.
However, when we ask what scientific breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
The success of the \textsl{AlphaFold} projects in protein structure \cite{alphafold} are often raised.
But these are successes in a very specific challenge-problem setting in which \emph{performance} is valued over \emph{understanding}.
In the natural sciences we almost exclusively care about understanding, in the long run.

@@

In contrast to natural science, ML research and ML methods are concerned with making accurate predictions for, or descriptions of, \emph{data}.
A ML method is considered successful if it performs well on held-out training data, even if the latent structure of the model is generic and the internals are impossible to interpret.
In ML, the considerations are all at the level of the data, and we are happy to use models where we have little or no understanding of the meanings or values of the latent parameters or weights.
Another way to ask our question is:
\emph{Where, in the natural sciences, can you use a model that you don't understand?}

In natural science the most important contributions and results are all at the level of the \emph{latents}:
We use data to learn about the latent structure of the world or of the system we are studying.
In astrophysics, this could be the interior structure of the Sun, or the statistics of planets around other stars, or the map of the dark matter surrounding the Milky Way.
The things we care about are almost never directly observable; they are parameters (or hyper-parameters) of a physical (or chemical or biological) model that predicts the observables.

@@

@@

\paragraph{Our contributions:}
\begin{itemize}
  \item We deliver a description of the fundamental \emph{ontology} and \emph{epistemology} of machine learning, and contrast these with the ontologies and epistemologies of the natural sciences.
  \item We elucidate two important and strong statistical biases that are being introduced to the natural sciences by some of the uses of ML in those disciplines. One is a confirmation bias that arises when simulations are replaced or augmented by emulators. Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint analyses. Neither of these two biases can be easily corrected.
  \item We show that there are many safe places for the use of ML methods in current natural-science practices, and places where (in the contemporary context) the use of ML methods is effectively \emph{required}. Many of these places are in the operational parts of scientific projects.
  \item We argue that in causal contexts, in which the ML method is used to model foregrounds, backgrounds, calibration parameters, or other confounders, using the most expressive ML method can deliver \emph{the most conservative approach} to the scientific problem.
\end{itemize}

@@

@@

For the purposes of this \documentname{} we will exclude more engineering-oriented questions such as: What protein sequences might be important for the treatment of diabetes? Or where in Canada might there be new uranium deposits? Or what kinds of catalytic surfaces might help turn biomass into biofuel?
This is a bit unfair!
We are excluding these kinds of questions in part because it is obvious that ML methods \emph{can} serve important purposes in such problems.
That will re-emerge as a theme, below.

@@

Unsupervised ML methods deliver descriptions or compressions of data.
Supervised ML methods find relationships between data (features) and data (labels).
In both cases, the methods are judged on their capability to accurately describe the data.
They are not judged on the details of their latent structure.
In a very important sense, the ML ontology is that \emph{only the data exist}.

In support of this point of view, here are some comments:
Deep-learning models have enormously large internal degeneracies (combinatorial degeneracies even).
These degeneracies are not seen as a problem, since all they do is make the latent weights less interpretable.
Contemporary optimization schemes are stochastic \cite{stochastic} and most models are non-convex.
The fact that it is impossible to find the global optimum---and the fact that in most contemporary models that isn't even the goal---shows that the latent parameters are not important.
The stabilization of optimizations with early stopping \cite{early_stop} and drop-outs \cite{dropout} demonstrate that there isn't even a goal of getting precisely to \emph{any optimum}.
Any practitioner using a 40-layer network can trivially (and without analysis) switch to a 42-layer, despite the fact that (from a functional point of view), this substantially changes the model capacity.

In ML, models are deemed stably optimized and useful not if the latent space---the joint values of all the weights and biases---is stable, but rather if the predictions for held-out data are stable.
Thus it is the data that exist in ML contexts, not the latent parameters.

In contrast, the ontologies of the natural sciences contain far more things than just the data.
In physics, for example, 
not only do the data exist, but so do forces, energies, momenta, charges, spacetime, wave functions, virtual particles, and much more.
These entities are judged to exist in part because they are involved in the latent structure of the successful theories; almost none of them are direct observables.
The most important discoveries in the natural sciences are discoveries of latent structure: Natural selection as an explanation for the differentiated properties of species \cite{natural_selection}, for example, or the elemental composition of the Sun \cite{sun_composition}, or the quarks and gluons that make up the proton \cite{proton_substructure}.

@@

Another critic might point to the literatures on interpretable ML \cite{interpretable} and explainable ML \cite{explainable}.
In these ML subfields, in some cases, attempts are made to understand the internal structure or the behavior of the ML model.
There are not enormous successes here, but if there were, these might be subfields in which the epistemology of ML gets more complex; maybe in the future some ML models will have to perform well on data \emph{and} be explainable in some important sense.
At present this isn't part of any ML epistemology.

In the natural sciences, in contrast, the epistemologies are much more restrictive and much more demanding than they are in ML.
The detailed epistemological framework depends on the natural-science field, and even on the practitioner within the field.
However, in all cases, it is more demainding:
A theory or explanation has to do much more than just explain the data in order to be accepted as true.
In physics for example, a model---which, as we note, is almost always a model of latent structure---is judged to be good or strongly confirmed not just if it explains observed data.
It must explain data in multiple domains, and it must also connect in natural ways to other theories or principles (such as conservation laws and invariances) that are strongly confirmed themselves.
(It is worthy of note here that a successful theory in the natural sciences is usually a combination of a more fundamental theory of the latents, and a less fundamental or auxiliary observation model, which explains how the latents appear in or affect the observable data.)
General relativity \cite{gr} was adopted not primarily because it explained anomalous data (although it did explain some); it was adopted because it had a beautiful structure, it resolved conceptual paradoxes in the pre-existing theory of gravity, and it was consistent with emerging ideas of field theory and geometry.

\section{Why do we need machine learning in the natural sciences?}\label{sec:why}
In this \sectionname{} we give some idea of how and why ML has had such a big impact in the natural sciences, and how and why many scientific projects require ML technologies.
In most cases, the ML is required on the engineering or execution side of projects that are large-scale in some sense.
\begin{description}
  \item[Label transfer:] Sometimes a project has informative data on a very large number of objects, but precise labels for only a few, maybe obtained through very careful analysis or external data.
  In this case, if it is extremely expensive to label more objects, a regression can be trained on the few labeled data points and then the trained regression used to label all the rest.
  Below (\secref{sec:bad}) we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used in downstream scientific projects.
  \item[Classification:] This is the same use case as label transfer, but in the specific case in which the labels are drawn from a finite (small, actually) set of discrete values.
  It is subject to the same kinds of biases described below (\secref{sec:bad}).
  \item[Speeding up decisions:] Many scientific projects must make decisions very fast in real time.
  The most extreme examples of this are in particle physics, where detectors (such as the \textsl{Atlas} Experiment \cite{atlas} at the \textsl{Large Hadron Collider}) must decide whether to trigger a data-saving event in a tiny fraction of second.
  It is often the case that trained ML classifiers can reproduce the selection boundaries as well---or nearly as well---as first-principles models, but with far less computation.
  It is critical, if ML methods are used for real-time operations, that the methods and all their latent parameters (weights) be preserved for analysis, operations simulations, and conterfactual exercises, often performed long after the data are taken.
  \item[Speeding up simulations:] In much of the natural sciences, the theory in play is a computation or simulation or digital twin.
  These simulations tend to be very computationally expensive, since they often span large ranges of spatial or time scales.
  ML regressions can be trained to emulate the simulations, or patch up low-resolution simulations to higher resolutions.
  Below (\secref{sec:bad}) we argue that the introduction of ML emulators can introduce an unwanted, strong confirmation bias.
  \item[Modeling nuisances:] In most natural-science domains, the quantities of interest are not directly observable, but rather the model or prediction for the data is a combination of the theory of relevance plus auxilliary theories of foregrounds, backgrounds, instrumentation, and noise sources.
  When the goal is for these nuisances to be effectively modeled but not necessarily understood in detail, ML approaches can be effective and (as we argue in \secref{sec:good}) even conservative.
  \item[Outlier detection:] In many of the sciences, discovery of new or previously unknown phenomena or objects can be of great importance.
  A recent example in astronomy is the discovery of fast radio bursts \cite{frbs}, which started as outliers in the data taken for imaging projects and have turned out to be interesting astrophysical objects.
  Since expressive unsupervised ML methods can describe accurately a complex distribution of data, they can also be used to find rare data points that are unlike any elements of the training data.
  \item[Information questions:] If a machine-learning method is correctly trained to predict labels $y$ from data $x$, and it succeeds (better than random, say), then it shows that data $x$ contains information within it about labels $y$.
  Regressions thus can be used to make one-sided measurements of quantities of information-theoretic interest.
  \item[Making discoveries?] There is a new hope in the sciences that sufficiently trained or constrained models might lead to new insights about scientific theories or might effectively make fundamental discoveries.
  There are approaches along the direction of symbolic regression \cite{symbolic1, symbolic2}, and there are approaches along the lines of foundation models \cite{foundation1}.
  So far this hope has borne no important new discoveries, but discoveries are not inconceivable.
  Indeed, there is a very real sense in which the motivations underlying foundation models---finding generally useful latent representations for qualitatively diverse data---are similar to the motivations underlying theoretical physics!
\end{description}
This list isn't exhaustive!
For example, it doesn't mention the use of ML to find summary statistics for the comparison of simulations and data in (say) simulation-based inferences \cite{sbi, simbig}.
It also doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, likelihood_ratio}.
The uses of ML in the natural sciences is legion.

@@

\paragraph{Amplifications of training-set biases:}
The outcomes of ML regressions are label estimates that are conditioned on the input features \emph{and also} on the totality of the training set used to train the regression.
This is good; the individual-data-point label estimates from the regression are the lowest ``risk'' (in the statistical sense) when they use non-zero bias in the bias--variance trade-off.
However, the biases that are weak or manageable on an individual data-point basis become strong or unacceptable biases when outputs of regressions are used jointly or in combination to measure a population or sub-population property.

This problem is demonstrated with a toy example in \appendixname~\ref{app:toy}.
It is worst when the regression is used to label a large data set or catalog or sample of data,
and when elements of that data set are used in populations or joint analyses.
In general, when multiple data point estimates are used jointly, the variance-induced offsets of the estimators average out but the bias-induced offsets remain fixed.
This is a straightforward point of statistics---this is not news---but it isn't currently informing most of the practice of ML in the natural sciences, where many catalogs and public data products are being produced with ML regressions.
% AFTER ICML submission: For example, HOGG CITES. HOGG: NOTE that we are self-citing here; we are guilty ourselves!

In principle, there are fixes for this problem that involve de-biasing the estimates.
This is not possible in general, because there is not usually enough validation data to accurately assess the bias.

>>>>

% To do:
% ----------------------------
% - Find the right place to cite the units paper.
% - audit bibliography; make sure arXiv references in the reference list are consistent? Also capitalization!
% - Update appendix with feedback from Vedant Chandra (email). In particular, modify model complexity and information content.
% - Should there be another BAD thing, which is about statistics for comparison of simulations and data? This is a great one; consider after ICML submission.
% - Publish on the arXiv.

% Style notes:
% ------------
% - We get an extra page for camera ready (9 pages + bibliography and appendices)
% - Be very careful about the SUBJECT of sentences. Is it ML that is bad? Is it the practice that is bad? Is it the practitioner that is bad?
% - Should we use NS for natural science?
% - Audit for flexible, expressive, high in capacity, and so on.
% - [more here]

% NOW it's GO TIME
\documentclass{article}

@@

\usepackage[accepted]{icml2024} %% HACK HACK

@@

% Hogg typesetting issues that won't conflict with ICML
\frenchspacing
\newcommand{\documentname}{\textsl{Position Paper}}
\renewcommand{\paragraph}[1]{\noindent\par\textbf{#1}}
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\newcommand{\figref}[1]{\figurename~\ref{#1}}
\newenvironment{hoggnumerate}
  {\begin{list}
    {\arabic{enumii}.}
    {\usecounter{enumii}
     %\setlength{\labelwidth}{3em}
     %\setlength{\labelsep}{0em}
     \setlength{\itemsep}{0in}
     \setlength{\parsep}{0in}
     \setlength{\parskip}{0in}
     %\setlength{\leftmargin}{1.5cm}
     %\setlength{\rightmargin}{2cm}
     %\setlength{\itemindent}{0em} 
     %\let\makelabel=\makeboxlabel
    }
  }
{\end{list}}

@@

\icmltitlerunning{Is machine learning good or bad for the natural sciences?}
\begin{document}

\twocolumn[
\icmltitle{\documentname: Is machine learning good or bad for the natural sciences?}

\begin{icmlauthorlist}
\icmlauthor{David W. Hogg}{ccpp,mpia,flatiron}
\icmlauthor{Soledad Villar}{flatiron,ams,minds}
\end{icmlauthorlist}
\icmlaffiliation{ccpp}{Center for Cosmology and Particle Physics, Department of Physics, New York University, USA}
\icmlaffiliation{mpia}{Max-Planck-Institut f{\"u}r Astronomie, Heidelberg, Germany}
\icmlaffiliation{flatiron}{Flatiron Institute, New~York, USA}
\icmlaffiliation{ams}{Department of Applied Mathematics and Statistics, Johns Hopkins University,  USA}
\icmlaffiliation{minds}{Mathematical Institute for Data Science, Johns Hopkins University, USA}
\icmlcorrespondingauthor{David W. Hogg}{david.hogg@nyu.edu}

@@

\vskip 0.3in
] % For the arXiv version, move this square bracket below the abstract

\printAffiliationsAndNotice{}
\begin{abstract}
  Machine learning (ML) methods are having a huge impact across all of the sciences.
  However, ML has a strong ontology---in which only the data exist---and a strong epistemology---in which a model is considered good if it performs well on held-out training data.
  These philosophies are in strong conflict with both standard practices and key philosophies in the natural sciences.
  Here we identify some locations for ML in the natural sciences at which the ontology and epistemology are valuable.
  For example, when an expressive machine-learning model is used in a causal inference to represent the effects of confounders, such as foregrounds, backgrounds, or instrument calibration parameters, the model capacity and loose philosophy of ML can make the results more trustworthy.
  We also show that there are contexts in which the introduction of ML introduces strong, unwanted statistical biases.
  For one, when ML models are used to emulate physical (or first-principles) simulations, they introduce strong confirmation biases.
  For another, when expressive regressions are used to label datasets, those labels cannot be used in downstream joint or ensemble analyses without taking on uncontrolled biases.
  The question in the title is being asked of all of the natural sciences; that is, we are calling on the scientific communities to take a step back and consider the role and value of ML in their fields; the (partial) answers we give here come from the particular perspective of physics.
\end{abstract}

\section{Introduction}\label{sec:intro}
It is an understatement to say that machine learning (ML) is having a big impact across the sciences.
A significant fraction of all scientific papers in the natural sciences now employ ML in part (or all) of their analyses.
(We will define ML below in \secref{sec:philosophy}).
However, when we ask what scientific breakthroughs have been enabled by this influx of new tools and methods, there isn't a long list.
The success of the \textsl{AlphaFold} projects in protein structure \cite{alphafold} are often raised.
But these are successes in a very specific challenge-problem setting in which \emph{performance} is valued over \emph{understanding}.
In the natural sciences we almost exclusively care about understanding, in the long run.

@@

In contrast to natural science, ML research and ML methods are concerned with making accurate predictions for, or descriptions of, \emph{data}.
A ML method is considered successful if it performs well on held-out training data, even if the latent structure of the model is generic and the internals are impossible to interpret.
In ML, the considerations are almost all at the level of the data, and we are happy to use models in which we have little or no understanding of the meanings or values of the latent parameters or weights.

In natural science, on the other hand, the most important contributions and results are all at the level of the \emph{latents}:
We use data to learn about the latent structure of the world or of the system we are studying.
In astrophysics, this could be the interior structure of the Sun, or the processes that form planets around other stars, or the map of the dark matter surrounding the Milky Way.
The things we care about are almost never directly observable; they are parameters (or hyper-parameters) of a physical (or chemical or biological) model that predicts the observables.
Often the thing we care about is the model itself.

@@

For another example, consider the discovery (by Kepler) that the paths of the planets are ellipses, with the Sun at one focus \cite{kepler}.
This discovery led to extremely precise predictions for data.
It was critical to this discovery that the data be well explained by the theory.
But that wasn’t the primary consideration that made the scientific community prefer the Keplerian model.
After all, the Ptolemaic model preceding Kepler made equally accurate predictions of held-out data.
Kepler’s model was preferred because it fit in with other ideas being developed at the same time, most notably heliocentrism \cite{copernicus} and Newtonian gravity \cite{newton}.

@@

\paragraph{Our contributions:}
\begin{itemize}
  \item We deliver a description of the fundamental \emph{ontology} and \emph{epistemology} of machine learning, and contrast these with the ontologies and epistemologies of the natural sciences.
  \item We elucidate two important and strong statistical biases that are being introduced to the natural sciences by some uses of ML. One is a confirmation bias that arises when simulations are replaced or augmented by emulators. Another is a more standard estimator bias that is (possibly enormously) amplified when elements of datasets produced by ML regressions are used in combination or joint or ensemble analyses. Neither of these two biases can be easily corrected.
  \item We show that there are many safe places for the use of ML methods in current natural-science practices, and places where (in the contemporary context) the use of ML methods is effectively \emph{required}. Many of these places are in the operational parts of scientific projects.
  \item We argue that in causal contexts, in which the ML method is used to model foregrounds, backgrounds, calibration parameters, or other confounders, using the most expressive ML method can deliver \emph{the most conservative approach} to the scientific problem.
\end{itemize}

@@

As broad as this definition is, it does not include standard model fitting or parameter estimation.
These tasks are usually performed in the context of a well-specified model, where the model does not qualitatively change as more data arrive, and the quantitative improvements to the parameter estimates only scale as something like the square root of the size of the data set.
Standard model fitting and parameter estimation are not ML in our sense.

@@

For the purposes of this \documentname{} we will exclude more engineering-oriented questions such as: What protein sequences might be important for the treatment of diabetes? Or where in Canada might there be new uranium deposits? Or what kinds of catalytic surfaces might help turn biomass into biofuel?
This is a bit unfair!
We are excluding these kinds of questions in part because it is obvious that ML methods \emph{can} serve important purposes in such problems.

In addition, we recognize that there are quite a few different attitudes that a scientist can take towards the concept of ``understanding''.
Does a program that can predict and control a robot arm accurately ``understand'' the arm?
Or does a system that can predict protein foldings ``understand'' protein folding?
We are using a restricted definition of ``understanding'' that excludes these operational successes; mostly in the natural sciences we are looking for semantic rather than operational understandings of systems.
But of course ML often \emph{can} provide extremely operationally capable systems.

@@

Unsupervised and self-supervised ML methods deliver representations, descriptions, or compressions of data.
Supervised ML methods find relationships between data (features) and data (labels).
In both cases, the methods are judged on their capability to accurately describe the data.
They are (often) not judged on the details of their latent structure.
In a very important sense, the ML ontology is that \emph{only the data exist}.

In support of this point of view, here are some comments:
Deep-learning models have enormously large internal degeneracies (combinatorial degeneracies even).
These degeneracies are not seen as a problem, since all they do is make the latent weights less interpretable without hurting performance \cite{belkin2019reconciling, bartlett2020benign}.
Contemporary optimization schemes are stochastic \cite{stochastic} and most models are non-convex.
The fact that it is impossible to find the global optimum---and the fact that in most contemporary models that isn't even the goal---shows that the latent parameters are not important.
The regularization of trained models with early stopping \cite{early_stop} and dropouts \cite{dropout} demonstrate that there isn't even a goal of getting precisely to \emph{any optimum}.
Any practitioner using a 23-layer network can trivially (and without analysis) switch to a 42-layer, despite the fact that (from a functional point of view), this might substantially change the model capacity or expressivity.

In ML, models are deemed stably optimized and useful not if the latent space---the joint values of all the weights and biases---is stable, but rather if the predictions for held-out data are stable.
This is used as an indicator that the \emph{learned function} (at least in the training, validation, and test set) is stable, even though there can be large subspaces of very different, unidentifiable latent parameters.
Since the learned function is a model for the data, or something that predicts the data, it is the data that exist in ML contexts, not the latent parameters.

In ML there is sometimes additionally a notion of ``ground truth''; does this concept extend the ontology of ML?
Some ML methods assume that the training (and validation) data constitute ground truth (the labels are correct and have knowledge of the world, say).
In this case, ``ground truth'' is just another name for the training data.
Other methods assume that the labels might be noisy or adversarial, and that the method discovers true or more true labels.
This does perhaps extend the ontology, but only slightly, since the ground truth is in the same space as the data (and nearby to it, presumably).
Another notion that sometimes arises in ML is the ``data-generating process'', especially in contexts of emulation.
We don't see this as different from the data themselves, but technically this does extend the ontology marginally.

In contrast to ML, the ontologies of the natural sciences contain far more things than just the data.
In physics, for example, 
not only do the data exist, but so do forces, energies, momenta, charges, spacetime, wave functions, virtual particles, and much more.
These entities are judged to exist in part because they are involved in the latent structure of the successful theories; almost none of them are direct observables.
The most important discoveries in the natural sciences are discoveries of latent structure: Natural selection as an explanation for the differentiated properties of species \cite{natural_selection}, for example, or the elemental composition of the Sun (see \citealt{sun_composition}), or the quarks and gluons that make up the proton \cite{proton_substructure}.

The mismatch of ontology (and epistemology, below) between ML and natural sciences is not unique to ML;
many tools used in the natural sciences have limited ontologies.
Perhaps the most extreme case is that of mathematical tools, which are based (in some sense) only on a limited set of axioms.
Mathematics has a very limited ontology.
ML is useful in science in a way that is analogous to other methodological tools; the point made here is that the particular and specific use of each tool ought to be matched to its ontology and epistemology.

@@

Another critic might point to the literatures on interpretable ML \cite{interpretable} and explainable ML \cite{explainable}.
In these ML subfields, in some cases, attempts are made to understand the internal structure or the behavior of the ML model.
There is not consensus here, but if there were, it might create subfields in which the epistemology of ML gets more complex; maybe in the future some ML models will have to perform well on data \emph{and} be explainable in some important sense.
That would have implications for the epistemology, and probably also the ontology of ML.

In the natural sciences, in contrast, the epistemologies are much more restrictive and much more demanding than they are in ML.
The detailed epistemological framework depends on the natural-science field, and even on the practitioner within the field.
However, in all cases, it is more demanding:
A theory or explanation has to do much more than just explain the data in order to be widely accepted as true.
In physics for example, a model---which, as we note, is almost always a model of latent structure---is judged to be good or strongly confirmed not only if it explains observed data.
It ought to explain data in multiple domains, and it must connect in natural ways to other theories or principles (such as conservation laws and invariances) that are strongly confirmed themselves.
(It is worthy of note here that a successful theory in the natural sciences is usually a combination of a more fundamental theory of the latents, and a less fundamental or auxiliary observation model, which explains how the latents appear in or affect the observable data.)
General relativity \cite{gr} was widely accepted by the community (and very quickly; \citealt{peebles}) not primarily because it explained anomalous data (although it did explain some); it was adopted because, in addition to explaining (a tiny bit of new) data, it also had good structure, it resolved conceptual paradoxes in the pre-existing theory of gravity, and it was consistent with emerging ideas of field theory and geometry.

\section{Why do we need machine learning in the natural sciences?}\label{sec:why}
In this \sectionname{} we give some idea of how and why ML has had such a big impact in the natural sciences, and how and why many scientific projects require ML technologies. 
In most cases, the ML is required on the engineering or execution side of projects that are large-scale in some sense.
The descriptions here are non-exhaustive, they have a strong physics bias to them, and they do not address the many technical nuisances that make this discussion approximate and less specific than we would like.
Our hope is that this \documentname{} is seen as a conversation starter and not as providing a complete answer to any of the posed questions.

\paragraph{Label transfer:}
Sometimes a project has informative data on a very large number of objects, but precise labels for only a few, maybe obtained through very careful analysis or external data.
In this case, if it is extremely expensive to label more objects, a regression can be trained on the few labeled data points and then the trained regression used to label all the rest.
Below in \secref{sec:bad} we argue that this is not an advisable use of ML in the natural sciences if the expectation is that the regression-generated labels are going to be used certain straightforward ways in downstream scientific projects.

\paragraph{Classification:}
This is the same use case as label transfer, but in the specific case in which the labels are drawn from a finite (small, actually) set of discrete values.
It is subject to the same kinds of issues as label transfer.

\paragraph{Speeding up decisions:}
Many scientific projects must make decisions very fast in real time.
The most extreme examples of this are in particle physics, where detectors (such as the \textsl{Atlas} Experiment \cite{atlas} at the \textsl{Large Hadron Collider}) must decide whether to trigger a data-saving event in a tiny fraction of second.
It is often the case that trained ML classifiers can reproduce the selection boundaries as well---or nearly as well---as first-principles models, but with far less computation.
It is critical, if ML methods are used for real-time operations, that the methods and all their latent parameters (weights) be preserved for analysis, operations simulations, and conterfactual exercises, often performed long after the data are taken.
% HOGG: KSF NOTES that this last point could be expanded upon here or in the discussion.

\paragraph{Speeding up simulations:}
In much of the natural sciences, the theory in play is a computation or simulation or digital twin.
These simulations tend to be very computationally expensive, since they often span large ranges of spatial or time scales.
ML regressions can be trained to emulate the simulations, or patch up low-resolution simulations to higher resolutions.
Below in \secref{sec:bad} we argue that the introduction of ML emulators can introduce an unwanted, strong confirmation bias.

\paragraph{Modeling nuisances:}
In most natural-science domains, the quantities of interest are not directly observable, but rather the model or prediction for the data is a combination of the theory of relevance plus auxilliary theories of foregrounds, backgrounds, instrumentation, and noise sources.
When the goal is for these nuisances to be effectively modeled but not necessarily understood in detail, ML approaches can be effective and, as we argue in \secref{sec:good}, even conservative.

\paragraph{Propose objects, materials, or interactions for follow-up:}
In most scientific settings, discoveries are valid if they can be experimentally verified, regardless of the process used to produce the scientific hypothesis.
That is, hypotheses can be generated by creative processes that don't need to be understood, provided that the hypotheses so generated can be tested by experimental methods that \emph{are} well understood.
ML-based generative models can be used to conjecture the existence of objects or materials with certain properties, that can be later verified in lab assays.
One example of this is ML-based drug discovery \cite{kang2018conditional}.

\paragraph{Outlier detection:}
In many of the sciences, discovery of new or previously unknown phenomena or objects can be of great importance.
A recent example in astronomy is the discovery of fast radio bursts \cite{frbs}, which started as outliers in the data taken for imaging projects and have turned out to be interesting astrophysical objects.
Since expressive unsupervised ML methods can (with some caveats) describe accurately a complex distribution of data, they can also be used to identify (some) rare data points that are unlike any elements of the training data.

\paragraph{Information questions:}
If a machine-learning method is correctly trained to predict labels $y$ from data $x$, and it succeeds (better than random, say), then it shows that data $x$ contains information within it about labels $y$.
Regressions thus can be used to make one-sided measurements of quantities of information-theoretic interest.

\paragraph{Making discoveries?}
There is a new hope in the sciences that sufficiently trained or constrained models might lead to new insights about scientific theories or might effectively make fundamental discoveries.
There are approaches along the direction of symbolic regression \cite{symbolic1, symbolic2}, and there are approaches along the lines of foundation models \cite{foundation1}.
So far this hope has borne no important new discoveries, but discoveries are not inconceivable.
Indeed, there is a very real sense in which the motivations underlying foundation models---finding generally useful latent representations for qualitatively diverse data---are similar to the motivations underlying theoretical physics.

\medskip
This list isn't exhaustive!
For example, it doesn't mention the use of ML to find summary statistics for the comparison of simulations and data in (say) simulation-based inferences \cite{sbi, simbig}.
It also doesn't mention using ML to emulate likelihoods or likelihood ratios \cite{biwei, likelihood_ratio}.
The uses of ML in the natural sciences are legion \cite{wang2023scientific,zhang2023artificial}.

@@

\paragraph{Amplifications of training-set biases:}
The outcomes of ML regressions are label estimates that are conditioned on the input features \emph{and also} on the totality of the training set used to train the regression.
This is good; the individual-data-point label estimates from the regression are the lowest ``risk'' (in the statistical sense) when they use non-zero bias in the bias--variance trade-off.
However, the biases that are weak or manageable on an individual data-point basis become strong or unacceptable biases when outputs of regressions are used jointly or in combination to measure a population or sub-population or ensemble property.

This problem is demonstrated with a toy example in \appendixname~\ref{app:toy}.
It is worst when the regression is used to label a large data set or catalog or sample of data,
and when elements of that data set are used in populations or joint or ensemble analyses.
In general, when multiple data point estimates are used jointly, the variance-induced offsets of the estimators average out but the bias-induced offsets remain fixed.
This is a straightforward point of statistics---this is not news---but it isn't currently informing most of the practice of ML in the natural sciences, where many catalogs and public data products are being produced with ML regressions.
% AFTER ICML submission: For example, HOGG CITES. HOGG: NOTE that we are self-citing here; we are guilty ourselves!

In principle, there might be fixes for this problem that involve de-biasing the estimates.
This is not possible in general, because there is not usually enough validation data to accurately assess the bias.

In Bayesian language, this problem is closely related to the point that when data are to be combined, they should be combined at the likelihood level, not the posterior level.
Likelihoods are not affected by population-level biases.
Likelihoods can produce unbiased point estimates.
Likelihood-based estimators are not usually the lowest-risk estimators, but they can be low- or zero-bias estimators.
The outputs of regressions are more like posterior-based estimators, affected by the implicit prior set by the training set.
When they are combined, the implicit prior gets amplified.
Most ML regressions are not capable of generating or emulating likelihoods.
There are interesting approaches in development to replace standard regressions with generative models that can produce likelihoods \cite{cannon, biwei, likelihood_ratio}.

Importantly, there is nothing unique to ML about low-risk estimators being biased, or that the use of biased estimators in populations analyses is problematic.
These are standard properties of estimators.
The unique thing to ML is that standard ML regressions almost always produce biased estimators.
Given the objective functions (cost functions) in use in ML regressions, this is a feature of ML, not a bug.
